### 2016-11-16 Manuscript/SW_early_life_telomeres.Rmd
Eds

#####Telomere dynamics and age
Using the cross-sectional data, we found that RTL decreased with age in the Seychelles warbler (LMM: estimate = `r round3(fixef(RTL_Age)['Agemonths'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Age)['Agemonths',],lessthan = F),collapse = ', ')`). RTL also decreased with log-transformed age (LMM: estimate = `r round2(fixef(RTL_LogAge)['LogAge'])`, CIs = `r paste(round2(CIwald(RTL_LogAge)['LogAge',]),collapse = ', ')`), and this model was a significantly better fit than one modelling a linear relationship between RTL and age (P = `r round3(anova(RTL_Age,RTL_LogAge)$"Pr(>Chisq)"[2])`), suggesting that telomere length decreased at the greatest rate in early life - somehting that was evident from the raw data (Fig. 1). All further models therefore used log-transformed age as a covariate. Also worth noting is that while the overall trend was for decreasing RTL, there were appoarent increases in cross-sectional telomere length after one year of age and at several points later in life (Fig. 1).

@@

#####Spatiotemporal variation in telomere dynamics and the environment
A linear mixed model of RTL in relation to social and ecological environmental variables revealed age, tarsus length, sex and annual variation in food availability were significantly related to RTL (Table 1). Model selection revealed that sex only appeared in models where tarsus length was also present (Table S1), and sex was no longer significant when tarsus length was removed from the full model (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`). RTL increased with tarsus length in males, but decreased in females. The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig. S1).

Model selection revealed that the top model explaining $\Delta$RTL was...

###Discussion
Here we use one of the largest datasets to date to assess the relationships between early-life conditions, telomere dynamics and survival in a closed population of Seychelles warblers. We found that, as in other studies, telomeres are lost at the greatest rate early in life. We found that telomere dynamics are subject to strong cohort effects. Within cohorts, we found little evidence that the ecological or social environment influenced telomere dynamics, and we found no evidence that early-life telomere length is related to juvenile or later-life survival.

Our study adds the the now substantial body of literature showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g. @Frenck1998; @Haussmann2003; @Heidinger2012; @Fairlie2016]. Our finding that tarsus length was related to juvenile RTL in an age-dependent manner (Fig. 3) most likely represents an additional age effect - in passerine birds tarsus length is correlated with age during the nestling stage, but not after fledging [@Ricklefs1976]. Telomere loss is most rapid early in life due to ongoing growth [@Frenck1998], and a negative correlation between RTL and body size in early life is therefore expected [see also @Ringsby2015]. We also found weak evidence for sex-specific telomere dynamics in Seychelles warblers, with males having longer-telomeres and higher loss of telomeres (Fig. 3C, Fig. 5B). We do not wish to speculate too much on this effect as confidence intervals for both sex effects overlapped zero. However, it is worth noting while we found no evidence that juvenile RTL has a sex-specific effect on survival, in adults the effect of telomere length on survival is strongest for males [@Barrett2013a]. Together, these results suggest that sex effects should at least be considered when studying any aspect of telomere dynamics [see also @Barrett2011].

 We found that at the individual-level, juvenile and adult telomere were not correlated, and that consistent telomere shortening only occurs  very early in life (Fig. 4). This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which juvenile and adult telomere length were highly correlated, and individual-level telomere shortening clearly occurs throughout the juvenile period, and into adulthood. However, in accordance with our study, a recent study on Soay sheep (*Ovis aries*) found only a very weak correlation between juvenile and adult RTL, and a highly complex pattern of telomere dynamics across the lifespan [@Fairlie2016]. It is too soon to speculate as to what may be generating this variation among species, but one possibility is that environmental variability itself affects the resolution with which we can observe telomere shortening in natural populations [see also @Fairlie2016]. Regardless, our data suggest the efficacy of telomeres as individual biomarkers of cost in wild populations will depend on the age structure of the study population, as well as the sampling resolution.

>>>>

#####Telomere dynamics and age
Using cross-sectional data we found that RTL decreased with age in the Seychelles warbler (LMM: estimate = `r round3(fixef(RTL_Age)['Agemonths'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Age)['Agemonths',],lessthan = F),collapse = ', ')`). RTL also decreased with log-transformed age (LMM: estimate = `r round2(fixef(RTL_LogAge)['LogAge'])`, CIs = `r paste(round2(CIwald(RTL_LogAge)['LogAge',]),collapse = ', ')`), and this model was a significantly better fit than one modelling a linear relationship between RTL and age (P = `r round3(anova(RTL_Age,RTL_LogAge)$"Pr(>Chisq)"[2])`), suggesting that telomere length decreased at the greatest rate in early life - somehting that was evident from the raw data (Fig. 1). All further models therefore used log-transformed age as a covariate. Also worth noting is that while the overall trend was for decreasing RTL, there were appoarent increases in cross-sectional telomere length after one year of age and at several points later in life (Fig. 1).

@@

#####Spatiotemporal variation in telomere dynamics and the environment
A linear mixed model of RTL in relation to social and ecological environmental variables revealed age, tarsus length, sex and annual variation in food availability were significantly related to RTL (Table 1). The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig. S1). One interesting finding from the model selection was that sex only appeared in top models where tarsus length was also present (Table S1). In accordance with this, when tarsus length was removed from the full model sex was no longer significant (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`), and a sex x tarsus interaction was significant when included (estimate = `r round3(fixef(RTL_Full_SexTarsus)['SexMales:Tarsus'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_SexTarsus)['SexMales:Tarsus',],lessthan = F),collapse = ', ')`); RTL increased with tarsus length in males, but decreased in females.

No social or ecological environmental variables were significant in a model of longitudinal data with $\Delta$RTL as the response variable (Table 2). The model averaging approach suggested that $\Delta$RTL was positively related to age (i.e. telomere shortening increased with age), and negatively associated with annual density (i.e. more shortening in years with higher population density; Table S2; Fig. S2).

###Discussion
Here we use a large, long-term, multi-cohort dataset to assess the relationships between the spatiotemporal variation in the ecological environment and lifelong telomere dynamics in a closed population of Seychelles warblers. We found that, as in other studies, telomeres shorten over time, and that telomere loss is greatest in early life. However, bouts of apparent telomere lengthening occur, and the extent of this lengthening cannot be explained solely by measurement error. We then found that telomere dynamics vary among sexes, and over time, with this temporal variation is related to fluctuations in food availability.

Our study adds the the now substantial body of literature showing that telomere length decreases with age in both humans and wild animals, and that this decrease is most rapid in early life [e.g. @Frenck1998; @Haussmann2003; @Heidinger2012; @Fairlie2016]. However, using longitudinal data we found that within-individual telomere measurments were only weakly correlated, and that consistent telomere shortening only occurs early in life, after which both shortening and lengthening occurs. This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which juvenile and adult telomere length were highly correlated, and individual-level telomere shortening clearly occurs throughout the juvenile period, and into adulthood. However, in accordance with our study, a recent study on a wild population of Soay sheep (*Ovis aries*) found only a very weak correlation between juvenile and adult RTL, and a highly complex pattern of telomere dynamics across the lifespan [@Fairlie2016].

We also found evidence for sex-specific telomere dynamics in Seychelles warblers, with males having longer-telomeres than females, in and that this sex-difference interacts with tarsus length. In Seychelles warblers the effect of telomere length on survival is strongest for males [@Barrett2013a]. Together, these results suggest that sex effects should be considered when studying any aspect of telomere dynamics [see also @Barrett2011].

### 2016-11-16 Manuscript/SW_early_life_telomeres_SOM.Rmd
Eds

FigS1 <- forestplot(table2)+
  annotate('text',x = 0.6,y = 0.05,label = 'A')+
  theme(axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black'))

@@

FigS2 <- forestplot(table2)+
  annotate('text',x = 0.6,y = 0.05,label = 'A')+
  theme(axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black'))

>>>>

FigS1 <- forestplot(table2)+
  theme(axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black'))

@@

FigS2 <- forestplot(table2)+
  theme(axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black'))
